---
title: "Stat 585 - Working with lists - Your Turn solution"
author: "Heike Hofmann"
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Your turn (10 mins)

- Extract other model diagnostics: find the R square value for each model. Draw a visualization of R square by country - can you see a pattern? What does the pattern mean?
- Extract residuals and fitted values for each of the models and store them in a dataset together with country and continent information. 
- Plot residuals across the years and fit a smooth. What does the pattern mean?
- Draw a scatterplot of predicted Life Expectancy in 2017 versus slope. Compare this to the scatterplot of Intercept versus slope.

```{r, echo=FALSE}
library(gapminder)
library(tidyverse)
gapminder <- gapminder %>% mutate(year = year-1950)

countryList <- gapminder %>% tidyr::nest(-country, -continent) %>% mutate(
  model = data %>% purrr::map(~lm(lifeExp~year, data = .))
)
```

## Extract R square for each model

```{r fig.height=4.5}
countryList <- countryList %>% unnest(model %>% purrr::map(broom::glance))
countryList$country <- reorder(countryList$country, countryList$r.squared)

countryList %>% 
  ggplot(aes(x = country, y = r.squared, colour=continent)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=-90, hjust=0)) +
  scale_colour_brewer(palette="Dark2")
```

## Extract residuals and fitted values

```{r}
countryFit <- countryList %>% select(country, continent, data, model) %>%
  mutate(fitted = model %>% purrr::map(broom::augment)) %>% 
  select(-model) %>% unnest()

countryFit
```

## Plot residuals across the years

```{r warning = FALSE, fig.height=3}
countryFit %>% 
  ggplot(aes(x = year, y = .resid)) +
  geom_line(aes(group = country, colour = continent)) +
  geom_smooth(method="loess") +
  facet_wrap(~continent)
```

## predicted Life Expectancy in 2017

```{r}
countryList <- countryList %>% 
  select(country, continent, data, model) %>%  
  mutate(
  le2017 = model %>% 
    purrr::map_dbl(function(m) predict(m, newdata=data.frame(year=67)))
)
countryList <- countryList %>% mutate(
  coefs = model %>% purrr::map(broom::tidy)
)
countryCoefs <- countryList %>% unnest(coefs) %>% select(country, continent, le2017, term, estimate) %>% spread(key=term, value=estimate)

```


## predicted Life Expectancy in 2017 and 1950

```{r fig.height=4, fig.width=5}
countryCoefs %>% 
  ggplot(aes(x = `(Intercept)`, y = le2017, colour=continent)) +
  geom_abline() +
  geom_point() + xlab("Predicted Life Expectancy in 1950") +
  ylab("Predicted Life Expectancy in 2017") +
  ggrepel::geom_label_repel(aes(label = country),
    data = countryCoefs %>% filter(`(Intercept)` > le2017))
```
