---
title: "Stat 585 - ggplot2"
author: "Heike Hofmann"
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# `ggplot2`

# Why `ggplot2`?

## Why `ggplot2`

- Wildly popular package for statistical graphics: over 1.75 million downloads from CRAN in 2015 (~ 6,200 times per day)
- Developed by Hadley Wickham (An ISU Alumni)
- Designed to adhere to good graphical practices
- Constructs plots using the concept of layers
- Supports a wide variety plot types and extensions
\medskip
- http://ggplot2.org/book/ or Hadley's book *ggplot2: Elegant Graphics for Data Analysis* for reference 

## Grammar of Graphics

A graphical representation (plot) consists of:

1. **mappings** (`aes`):  data variables are mapped to graphical elements
2. **layers**:  geometric elements (`geoms`, such as points, lines, rectangles, text, ...) and statistical transformations (`stats`, are identity, counts, bins, ...)
3. **scales**: map values in the data space to values in an aesthetic space (e.g. color, size, shape, but also position)
4. **coordinate system** (`coord`): normally Cartesian, but pie charts use e.g. polar coordinates
5. **facetting**: for small multiples (subsets) and their arrangement
6. **theme**: fine-tune display items, such as font and its size, color of background, margins, ...

## Mapping Data

```{r echo = FALSE}
FortDodge <- readRDS("../data/USH00132999.rds")
library(ggplot2)
```

```{r fig.width = 6, fig.height = 3, warning = FALSE}
ggplot(data = FortDodge, aes(x = Year+Month/12, Temp_Avg/100)) + 
  geom_point() + 
  xlim(c(2011,2016))
```

## Changing geometric elements

```{r fig.width = 6, fig.height = 3, warning = FALSE}
ggplot(data = FortDodge, aes(x = Year+Month/12, Temp_Avg/100)) + 
  geom_line() + 
  xlim(c(2011,2016))
```

## Adding data layer

```{r fig.width = 6, fig.height = 3, warning = FALSE}
ggplot(data = FortDodge, aes(x = Year+Month/12, Temp_Avg/100)) + 
  geom_line() + geom_point() + 
  xlim(c(2011,2016))
```

## Adding summary layer

```{r fig.width = 6, fig.height = 3, warning = FALSE}
ggplot(data = FortDodge, aes(x = Year+Month/12, Temp_Avg/100)) + 
  geom_point() + geom_smooth(method = "lm") +
  xlim(c(2011,2016))
```

## Theming

```{r fig.width = 6, fig.height = 3, warning = FALSE}
ggplot(data = FortDodge, aes(x = Year+Month/12, Temp_Avg/100)) + 
  geom_point() + geom_smooth(method = "lm") +
  xlim(c(2011,2016)) + 
  xlab("Time") + ylab("Average Temperature") +
  xkcd::theme_xkcd()
```

## `ggplot2` resources

Generally in `ggplot2`: every aspect of a plot can be changed, but there are good defaults based on cognitive principles.

RStudio's [cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)

Online [`ggplot2` documentation](http://docs.ggplot2.org/current/)

Email list (sign up from http://ggplot2.org/)

## Your Turn (10 min)

Load the temperature data [temps.rds](https://github.com/Stat585-at-ISU/materials/blob/master/data/temps.rds) with `readRDS`. 

1. Pick one weather station and plot the monthly temperatures using the `ggplot2` package. Use a line to show the time series.
2. Now expand this plot to show information for two weather stations. 
3. Label the time series by the station ID.

## Layers

each layer has several parts, the two most important are:

- mapping: using the `aes()` function, we specify mappings between variables and aesthetics of the chart
- `data`: the dataset used in the layer 

`ggplot` is the layer that sets the defaults, <br>
`geom_XXX` creates a layer: for each aspect it either uses the default or has to specify its own

## Default versus Layer Specification

```{r, fig.width=9.5, fig.height = 3, message = FALSE}
library(tidyverse)
temps <- readRDS("../data/temps.rds")
temps %>% filter(Station %in% c("USH00012813", "USH00454764")) %>%
  ggplot(aes(x = Year+Month/12, y = Temp)) +
  geom_line(aes(group = Station, colour=Station)) +
  geom_text(data = temps %>%
              filter(Year == 2016, Month==7,
                     Station %in% c("USH00012813", "USH00454764")),
            hjust = 0,
            aes(label = Station)) + xlim(c(1950, 2020))
```

##

```{r, fig.width=9.5, fig.height = 3, message = FALSE}
temps %>% filter(Station %in% c("USH00012813", "USH00454764")) %>%
  ggplot(aes(x = Year+Month/12, y = Temp, colour=Station)) +
  geom_line(aes(group = Station)) +
  geom_text(data = temps %>%
              filter(Year == 2016, Month==7,
                     Station %in% c("USH00012813", "USH00454764")),
            hjust = 0,
            aes(label = Station)) + xlim(c(1950, 2020))
```
