---
title: "Stat 585 - Databases in R"
author: "Heike Hofmann"
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---
```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Databases and R

## Outline

- a bit about databases
- linking R to a (local) data base
- investigate the FARS data

<br>

Resources:

- `vignette("databases", package="dplyr")` (for working with dplyr and databases)
- `vignette("two-table", package="dplyr")` (for joining tables - we've covered this before)


## Why databases?

What is a database?

- A collection of data
- A set of rules to manipulate data

Why are they important?

- Efficient manipulation of large data sets
- Convenient processing of data
- Integration of multiple sources of data
- Access to a shared resource


## Relational Databases

- Database is collection of tables and links (normal form)
- SQL (Structured Query Language) for querying 
- DBMS (Database Management System) and managing data

## Structured Query Language

* Structured Query Language (1970, E Codds)
* Programming language used for accessing data in a database
* ANSI standard since 1986, ISO standard since 1987
* Still some portability issues between software systems!
* We’ll mainly focus on SQL queries to access data

## Syntax

* SQL is not case sensitive.
* Some systems require ```;``` at the end of each line. The semi-colon can be used to separate each SQL statement in a system that allows multiple command to be executed in a call to the server.


## Select

<div class="double">
<p class="double-flow">
**Student**

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |

</p>
<p class="double-flow">
**Attendance**

| ***ID***            | ***Date***          | Status  |
| ------------- |:-------------:| -------:|
| 1234          | 02-Feb        | absent  |
| 1234          | 09-Feb        | absent  |
| 4321          | 02-Feb        | present |
| 4321          | 09-Feb        | absent  |

</p>
</div>


```Select * from student``` leads to 

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |






## Where


<div class="double">
<p class="double-flow">
**Student**

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |

</p>
<p class="double-flow">
**Attendance**

| ***ID***            | ***Date***          | Status  |
| ------------- |:-------------:| -------:|
| 1234          | 02-Feb        | absent  |
| 1234          | 09-Feb        | absent  |
| 4321          | 02-Feb        | present |
| 4321          | 09-Feb        | absent  |

</p>
</div>

```Select Name from student where Major='Math'``` leads to 

| Name          | 
| ------------- |
| Never Ever    | 


## Aggregations

<div class="double">
<p class="double-flow">
**Student**

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |

</p>
<p class="double-flow">
**Attendance**

| ***ID***            | ***Date***          | Status  |
| ------------- |:-------------:| -------:|
| 1234          | 02-Feb        | absent  |
| 1234          | 09-Feb        | absent  |
| 4321          | 02-Feb        | present |
| 4321          | 09-Feb        | absent  |

</p>
</div>

```Select ID, count(ID) from Attendance where Status='absent'``` leads to 

| ***ID***            | Frequency          | 
| ------------- |-------------:| 
| 1234          | 2    | 
| 4321          | 1    | 


## Additional Resources for SQL

* W3Schools: http://www.w3schools.com/sql/default.asp
* DBI package, RMySQL, RSQLite: 
http://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf
http://cran.r-project.org/web/packages/RSQLite/RSQLite.pdf

## Good news!

- `dplyr` package works (almost) the same for local data frames as tables in a database 
- `dplyr` functionality:
`group_by`, `summarize`, `transform`, `filter`, `arrange`, `select`

## An example: FARS data

- US Department of transportation is keeping a record of every accident that results in a fatality in the FARS Data base (fatal accident report system, http://www.nhtsa.gov/FARS)
- FARS consists of 20+ tables consisting of various aspects of each accident
Documentation at
- http://www-nrd.nhtsa.dot.gov/Cats/listpublications.aspx?Id=J&ShowBy=DocType

## SQLite

- SQLite is a relational database management system
- Unlike other DBMS SQLite is a server-less system with "zero-configuration"
- `fars2014` is a SQLite database of the 2014 data provided by FARS, consisting of three tables: `accidents`, `person`, and `vehicle` (make sure to download  `fars2014`  from our github repo).


connecting to the database:

```{r}
library(dplyr)
fars <- src_sqlite("../data/fars2014")
fars
```

## Connecting to FARS

Use `tbl` to connect to a specific table (dataset) in the database

note: R does not load the data into the session

```{r}
accidents <- tbl(fars, "accidents")
accidents
```

## 

- `dplyr` commands do not actually download the (whole) database
- use `collect()` to extract *all* records. This returns a tibble

```{r eval = FALSE}
accidents %>% 
  filter(between(LONGITUD, -130, 0)) %>% 
  collect() %>%
  ggplot(aes(LONGITUD, LATITUDE)) +
    geom_point(alpha = 0.5, size = 0.5)
```



##

```{r echo=FALSE}
accidents %>% 
  filter(between(LONGITUD, -130, 0)) %>% 
  collect() %>%
  ggplot(aes(LONGITUD, LATITUDE)) +
    geom_point(alpha = 0.5, size = 0.5)
```

## Your Turn (8 mins)

Connect to the FARS database (using the SQLite database `fars2014`). 
Answer the following questions:

- are there some days of the week where more accidents happen than on others (use variable `DAY_WEEK`)?
- what time of the day do accidents happen (use variable `HOUR`)?
- what is the number of accidents with at least one drunk driver (use variable `DRUNK_DR`)?

<br>
<br>
<br>
<br>

```{r}
dayweek <- accidents %>% group_by(DAY_WEEK) %>% tally() 
dayweek %>% collect() %>% 
  ggplot(aes(x = DAY_WEEK, weight=n)) + geom_bar() 

hours <- accidents %>% group_by(HOUR) %>% tally() 
hours %>% filter(HOUR < 25) %>% collect() %>% 
  ggplot(aes(x = HOUR, weight=n)) + geom_bar() 

drunk <- accidents %>% group_by(DRUNK_DR) %>% tally() 
drunk %>% collect() %>% 
  ggplot(aes(x = DRUNK_DR, weight=n)) + geom_bar() 
```

## Your Turn (5 mins)

Connect to the  `person` table. Identify drivers (`PER_TYP` == 1, see [fars manual](https://www-fars.nhtsa.dot.gov/Common/IdDefs.aspx) ) and subset on them.

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
person <- tbl(fars, "person")
drivers <- person %>% filter(PER_TYP == 1)
```

## Working with multiple tables

- Only rarely single datasets are giving us all the answers we need
- `left_join`, `right_join` work in the same way on databases as on local data frames.


## Investigating accidents by gender

We want to join drivers and accidents:

```{r}
driver_acc <- left_join(drivers, accidents)
```

Note that there are a lot of variables with the same name - we actually include all of these variables 


## Your Turn (5 min)

- Join drivers and accidents data.
- Tally the number of accidents by day of the week (`DAY_WEEK`), hour of the day (`HOUR`) and gender (`SEX`). Visualize the results!


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}
driver_acc %>% group_by(DAY_WEEK, HOUR, SEX) %>% 
  tally() %>% 
  filter(HOUR < 25, SEX < 8) %>% 
  collect() %>% 
  ggplot(aes(x = HOUR, y = n)) + 
    geom_point(aes(colour = factor(SEX))) +
  facet_wrap(~DAY_WEEK) +
  scale_colour_brewer(palette="Set1")
```
