---
title: 'Lab Assignment #5'
author: "Heike Hofmann"
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab #5: The making of an R package

## Lab Assignment

The objective of today's lab is to create and R package, document it and make it available through a github repository.

1. Create an R package with at least one (exported) function and one dataset. 
2. Document both the function and the dataset appropriately. 
3. Place your R package on github, such that it is possible to install via the `install_github` function in the `devtools` package (e.g. try `install_github("heike/ggmapr")`).
4. Write a `README.md` file to showcase your package.
5. Make sure that your package passes CRAN check, i.e. run `devtools::check()` without any warnings or errors.

## Your Solutions

```{r, eval=FALSE}
library(devtools)
install_github("bbuzzee/package2")
library(help=package2)
library(package2)
?mod
mod(5,3)
```
The other packages: `joepapio/585lab5`, `calegacy/Lab5PackageR`, `mstuart2097/Lab5`, `jeffsong9/FREDdata`, `akshayayadav/Team8`, `KatieWill/LittleChickens`, `haleyjeppson/cousinR`
