---
title: "585Lab3"
date: "February 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(purrr)
```

## Data Cleaning

First step, read in the files. 


```{r}
file1 <- lapply(excel_sheets("spreadsheets/FileOne.xlsx"), read_excel, path = "spreadsheets/FileOne.xlsx")
file2 <- lapply(excel_sheets("spreadsheets/FileTwo.xlsx"), read_excel, path = "spreadsheets/FileTwo.xlsx")

```

Before we can combine the files, or even the semesters, we need to do a little bit of cleaning.

```{r }
# First up, file 1:

file1[[1]]
# file 1 sem 1 is missing column titles
names(file1[[1]])[c(1,2)]<-c("ID", "Test")
file1[[1]] <- file1[[1]] %>% mutate ( Sem =1)


file1[[2]]
# file 1 sem 2 is also missing column titles
names(file1[[2]])[c(1,2)]<-c("ID", "Test")
# delete extra column of NA's
file1[[2]] <- file1[[2]][,-15]
file1[[2]] <- file1[[2]] %>% mutate ( Sem = 2)

file1[[3]]
# file 1 sem 3 is also missing column titles
names(file1[[3]])[c(1,2)]<-c("ID", "Test")
file1[[3]] <- file1[[3]] %>% mutate ( Sem =3)

file1[[4]]
# file 1 sem 4 is also missing column titles
names(file1[[4]])[c(1,2)]<-c("ID", "Test")
# the ids for sem 4 have an extra 100 in them?
file1[[4]] <- file1[[4]] %>% mutate ( ID=ID -100, Sem =4)

# combine the list of data frames into one dataframe
f1 <- do.call("rbind", file1)

# make sure all "test" resposes are the same case
f1$Test <- gsub("PRE", "Pre", f1$Test)
f1$Test <- gsub("POST", "Post", f1$Test)
f1

#fix names of the columns
names(f1)[c(14, 15,18)] <- c("Normalized Change", "Gender", "Treatment Part 2")
f1


```

```{r }
# Next up, file 2:

file2[[1]]
# file 1 sem 1 is missing column titles
names(file2[[1]])[c(1,2)]<-c("ID", "Test")
file2[[1]] <- file2[[1]] %>% mutate (Sem =1)


file2[[2]]
# file 2 sem 2 is also missing column titles
names(file2[[2]])[c(1,2)]<-c("ID", "Test")
file2[[2]] <- file2[[2]] %>% mutate ( Sem =2)

file2[[3]]
# file 2 sem 3 is also missing column titles
names(file2[[3]])[c(1,2)]<-c("ID", "Test")
file2[[3]] <- file2[[3]] %>% mutate ( Sem =3)

file2[[4]]
# file 1 sem 4 is also missing column titles
names(file2[[4]])[c(1,2)]<-c("ID", "Test")
# the ids for sem 4 have an extra 1100 in them?
file2[[4]] <- file2[[4]] %>% mutate ( ID= ID -1100, Sem =4)

# looking back throu the dimensions of all 4 semesters, semester one is missing the column "MTH 3", so we'll add in a column of NAS
file2[[1]] <- file2[[1]] %>% mutate ("MTH 3" = NA)
#but I would like the columns to still be in the same order, so I will rearrange them
file2[[1]]  <- file2[[1]][,c(1:16,41,17:40)]

# combine the list of data frames into one dataframe
f2 <- do.call("rbind", file2)

# make sure all "test" resposes are the same case
f2$Test <- gsub("PRE", "Pre", f2$Test)
f2$Test <- gsub("POST", "Post", f2$Test)
f2 %>% arrange(ID)
names(f2)[37]<- "Gender"


```

Now that we have the two files in two dataframes we can start working on getting them combined into one dataframe.

We have a split key so we will need to create two dataframes for each file before continuing.

First, we should extract the all the information about the student that is constant throughout the semester.

```{r}

keys1 <- f1 %>% select(ID, Gender, Characteristic, `Treatment Part 1`, `Treatment Part 2`, Sem)
keys2 <- f2 %>% select(ID, Gender, Characteristic, `Treatment Part 1`, `Treatment Part 2`, Sem)
#combine the two together
keys <- rbind(keys1, keys2)



#since there are duplicates, we want to pull out the unique, complete, values
keys <- unique(keys)
keys <- keys[complete.cases(keys),]
keys %>% arrange(ID)
```


Now we can join the f1 and f2 with our dataframe of keys

```{r}
#take gender and characeristic out of files 1 and 2
F1 <- f1[,-c(15:19)]
F2 <- f2[,-c(37:41)]

# Join the dataframes
first <- left_join(keys, F1, by=("ID"))
first %>% arrange(ID)
final <- left_join(first, F2, by=c("ID","Test"))
final %>% arrange(ID)

```


After joining the two dataframes, we renamed the variables corresponding to each test total score.


```{r}

# Renaming variables
names(final)[6] <- "Semester"
names(final)[18] <- "Test1_Total"
names(final)[52] <- "Test2_Total"
```

## Graphs Comparing Pre and Post Scores


Examining the number of IDs and unique IDs in the key dataframe created earlier, the lengths differ, due to discrepancies between the test 1 and test 2 variables for characteristic and gender. There are several examples below of students that show these discrepancies.

```{r}
# Comparing lengths of key IDs and unique IDs
length(keys$ID)
length(unique(keys$ID))

# Example of disagreement in characteristic value
final %>%
  filter(ID == "30033") %>%
  select(ID, Characteristic)

# Example of disagreement in gender value
final %>%
  filter(ID == "40030") %>%
  select(ID, Gender)
```

The discrepancies from test 2 dataframe occur at the end of the final dataframe. Although it would be advised to consult with the data collaborator, the duplicates after the first 346 rows are removed on the assumption that the values from the test 1 dataframe are correct .

```{r}
# Removing rows with dis
final <- final %>%
  slice(1:346)
```

With these rows removed, the dataframe can be transformed into the necessary formats using functions from the *tidyr* package. The test 1 and test 2 scores are separated into two dataframes with pre and post scores in two columns. Only the variables of ID, Gender, Test, and Semester are kept. Additionally, we created a variable of the difference between the post and pre scores (Post-Pre).

To get a better feel for differences between Pre and Posttest scores, as well as differences by gender, several ways of visually comparing the pre and post scores are included below. For each test, we produce a scatterplot comparing Pre and Post scores, then side by side box plots of Pre and Post differences by gender, and lastly overlapping densities of the differences between Pre and Post by gender.

```{r}
# Data for test 1 scores
graph.data.1 <- final %>%
  select(ID, Gender, Test, Test1_Total, Semester) %>%
  spread(key = Test, value = Test1_Total) %>%
  mutate(Difference = Post - Pre)

# Data for test 2 scores
graph.data.2 <- final %>%
  select(ID, Gender, Test, Test2_Total, Semester) %>%
  spread(key = Test, value = Test2_Total) %>%
  mutate(Difference = Post - Pre)
```

### Test 1 Scores

```{r, warning = FALSE}
ggplot(graph.data.1, aes(x = Pre, y = Post, color = Gender)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1) +
  facet_wrap( ~ Semester) +
  labs(title= "Scatterplot of Pre and Post Test 1 scores (black line is the 1:1 line)")

ggplot(graph.data.1, aes(x = Gender, y = Difference)) + 
  geom_boxplot() +
  facet_wrap( ~ Semester) +
  coord_flip() +
  labs(title= "Side by side boxplots of differences between Pre and Post Test 1 scores")

ggplot(graph.data.1, aes(x = Difference)) + 
  geom_density(data = subset(graph.data.1, Gender == 'Female'),fill = "red", alpha = 0.5) +
  geom_density(data = subset(graph.data.1, Gender == 'Male'),fill = "blue", alpha = 0.5) +
  facet_wrap( ~ Semester) +
  labs(title= "Overlapping densities of differences between Pre and Post Test 1 scores")
```

### Test 2 Scores

```{r, warning = FALSE}
ggplot(graph.data.2, aes(x = Pre, y = Post, color = Gender)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1) +
  facet_wrap( ~ Semester) +
  labs(title= "Scatterplot of Pre and Post Test 2 scores (black line is the 1:1 line)")

ggplot(graph.data.2, aes(x = Gender, y = Difference)) + 
  geom_boxplot() +
  facet_wrap( ~ Semester) +
  coord_flip()+
  labs(title= "Side by side boxplots of differences between Pre and Post Test 2 scores")

ggplot(graph.data.2, aes(x = Difference)) + 
  geom_density(data = subset(graph.data.2, Gender == 'Female'),fill = "red", alpha = 0.5) +
  geom_density(data = subset(graph.data.2, Gender == 'Male'),fill = "blue", alpha = 0.5) +
  facet_wrap( ~ Semester) +
  labs(title= "Overlapping densities of differences between Pre and Post Test 2 scores")
```





