---
title: "Stat 585 - Crosstalk"
author: "Heike Hofmann"
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---
```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Making widgets talk to each other

## What is `crosstalk`?

- `crosstalk` is an add-on R package to `htmlwidgets` produced by RStudio.
- extends htmlwidgets and allows implementing cross-widget interactions: linked brushing and filtering. <br><br><br><br><br><br>
- Tutorial: http://rstudio.github.io/crosstalk/

## crosstalk Example

[Result](03_crosstalk-example-01.html)

```{r, eval = FALSE, fig.width=5, fig.height=4}
library(crosstalk)
library(shiny)
library(d3scatter)

shared_mtcars <- SharedData$new(mtcars)
bscols(widths = c(3,NA,NA),
  list(
    filter_checkbox("cyl", "Cylinders", shared_mtcars, 
                    ~cyl, inline = TRUE),
    filter_slider("hp", "Horsepower", shared_mtcars, 
                  ~hp, width = "100%"),
    filter_select("auto", "Automatic", shared_mtcars, 
                  ~ifelse(am == 0, "Yes", "No"))
  ),
  d3scatter(shared_mtcars, ~wt, ~mpg, ~factor(cyl), 
            width="300", height=250),
  d3scatter(shared_mtcars, ~hp, ~qsec, ~factor(cyl), 
            width="300", height=250)
)
```

## Another - crazy(!) - example

[Result](03_crosstalk-example-02.html)

```{r, eval = FALSE, warning = FALSE, fig.width=5, fig.height=4}
library(gapminder)
library(crosstalk)
library(ggplot2)
library(plotly)

g <- crosstalk::SharedData$new(gapminder, ~continent)

gg <- ggplot(g, aes(gdpPercap, lifeExp, color = continent, 
                    frame = year)) +
  geom_point(aes(size = pop, ids = country)) +
  geom_smooth(se = FALSE, method = "lm") +
  scale_x_log10()

ggplotly(gg) %>% 
  highlight("plotly_hover")
```


## Sharing data between charts

**`SharedData`**: puts the data frame into a reactive expression, adds a key and a crosstalk group (crosstalk identifier, different objects can have the same crosstalk group). 

```{r warning=FALSE, message=FALSE}
library(crosstalk)
shared_iris <- SharedData$new(iris)
shared_iris$key() # row names are used by default
```

## Filter functions

- `filter_checkbox`, `filter_slider`, `filter_select` are crosstalk input function
- similar to `shiny` and `ggvis`
- Input data must be a SharedData object, filter selections are stored and visualizations updated correspondingly


## Linked brushing

- Function `bscols` helps with the layout in the webpage
- Brushing is enabled in `d3scatter` plots (like selection filter)
- Corresponding points are shown in **linked** plots (i.e. plots using the same data)

[Result](03_crosstalk-example-03.html)

```{r, eval = FALSE}
library(crosstalk)
library(leaflet)

shared_quakes <- SharedData$new(quakes[sample(nrow(quakes), 100),])
bscols(
  leaflet(shared_quakes, width = "100%", height = 300) %>%
    addTiles() %>%
    addMarkers(),
  d3scatter(shared_quakes, ~depth, ~mag, width = "100%", height = 300)
)
```


## Something doesn't work?

- Drawback of the bleeding edge: packages are still under active development
- Package versions are important

```{r}
devtools::session_info()
```


## Your Turn (5 min)

- Copy the code for the `mtcars` example into a new RMarkdown file and run it.
- Add a scatterplot of displacement (`disp`) and horse power (`hp`)
- Add an input filter for the number of carburetors.

## Your Turn (10 min)

- Exchange on of the scatterplots for a `plotly` chart. What happens with the linked brushing and filters?

## ggplot2 and crosstalk

[Result](03_crosstalk-example-04.html)

```{r, eval = FALSE}
sd <- SharedData$new(mtcars)
bscols(  
filter_checkbox("cyl", "Cylinders", sd, 
                    ~cyl, inline = TRUE),
    filter_slider("hp", "Horsepower", sd, 
                  ~hp, width = "100%"),
    filter_select("auto", "Automatic", sd, 
                  ~ifelse(am == 0, "Yes", "No"))

)
renderPlot({
   df <- sd$data(withSelection = TRUE, withFilter = TRUE)
   ggplot(df, aes(wt, mpg,
     color = factor(cyl))) +
     geom_point() 
   })
```
