---
title: "Stat 585 - Profiling"
author: "Heike Hofmann"
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---
```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rmarkdown)
```

# Writing good code

## What makes code "good"?

- maintainable
- extendable
- fast

How do we recognize good code? How do we spot bad code?

## Getting better

- Read good code, try to emulate
- READ code 
- spot GOOD code - how?
- Re-read your own (old) code CRITICALLY. 
- Possible questions:  how can I break the code?  how can I write this shorter? how can I make this more general?
- Foremost:  PRACTICE!


## Read code


What does this function do?
How does it work?

```{r eval = FALSE}
somefunction <- function (...) 
{
    k <- length(ll <- list(...))
    if (k == 0L) 
        return(invisible())
    mc <- match.call()
    for (i in 1L:k) if (!(is.logical(r <- ll[[i]]) && !any(is.na(r)) && 
        all(r))) {
        ch <- deparse(mc[[i + 1]], width.cutoff = 60L)
        if (length(ch) > 1L) 
            ch <- paste(ch[1L], "....")
        stop(paste(ch, " is not ", if (length(r) > 1L) 
            "all ", "TRUE", sep = ""), call. = FALSE)
    }
    invisible()
}
```

## Step-by-step code checking

The ```browser()``` function allows stepping through your code.

```{r}
help(browser)
```

- ```n```= executes the next line of code
- ```c```= continue execution until the end of the function, and exit
- ```where``` gives you the stack information, the trace of the active function calls
- ```Q``` exits browser()

## Simple alternatives

Putting in 

- ```cat()```
- ```str()```
- ```head()```
- ```summary()```
- ```print()```

are simple ways to build code up step-wise ensuring that each piece works


## Let's try browser()

```somefunction2()``` function has ```browser()``` in first line of function

```{r echo = FALSE}
somefunction2 <- function (...) 
{
    browser()
    k <- length(ll <- list(...))
    if (k == 0L) 
        return(invisible())
    mc <- match.call()
    for (i in 1L:k) if (!(is.logical(r <- ll[[i]]) && !any(is.na(r)) && 
        all(r))) {
        ch <- deparse(mc[[i + 1]], width.cutoff = 60L)
        if (length(ch) > 1L) 
            ch <- paste(ch[1L], "....")
        stop(paste(ch, " is not ", if (length(r) > 1L) 
            "all ", "TRUE", sep = ""), call. = FALSE)
    }
    invisible()
}
```

```{r eval = FALSE}
x <- 1
somefunction2(x==x, 1+1==2, c(7+5==11.99999, 1+1==2))
```

Browser enables:

- printing of result of each line
- checking changes to values
- make sure that it results in what is expected

## `debug()`

```debug(f)``` automatically places a ```browser()``` statement in the first line of function `f`

```undebug(f)``` removes it.

## Setting error handling

- ```options(error=recover)``` will drop user into ```browser()``` upon an error
- ```options(error=NULL)``` default, do nothing
- ```options(warn=2)``` sets maximum number of warnings to be 2
- ```options(warn=0)``` warnings are stored until top level function finishes
- ```options(warn=-1)``` ignore all warnings

## Traceback

```traceback()``` shows the stack at the time the last error occurred
```{r eval=FALSE}
f <- function(x) { 
  w(x)
	g(h(x)) 
	w(x)
} 
g <- function(x) {
	a <- 10
	x
} 

h <- function(x) {
	w(x) 
	w(x) 
}
	 
w <- function(x) { 
	if (sample(10, 1) == 1) stop("This is an error!")
}

f() 
traceback()
```

## Traceback

  
```
> f()
Error in w(x) : This is an error! 
> traceback()
5: stop("This is an error!") at #2
4: w(x) at #2
3: h(x) at #3
2: g(h(x)) at #3
1: f()
```

## Your Turn: fix this!! (8 mins)

```larger(x,y)``` is supposed to return the element-wise maximum of two vectors

```{r}
larger <- function(x, y) { 
  y.is.bigger <- y > x 
	x[y.is.bigger] <- y[y.is.bigger] 
	x
} 
larger(c(1, 5, 10), c(2, 4, 11)) 
larger(c(1, 5, 10), 6)
```


## Signs of inexperienced programmers


- Violation of the language model
    - e.g. R is vector oriented language, i.e.  ```for``` loops should be used sparingly, 
    - e.g. use of ```rbind``` (it is *known* to be very slow)
- Insufficient knowledge of the language. We're all guilty of that at different levels 
- Reading other people's code helps improve our language knowledge!

## Removing for loops

```{r}
x <- floor(runif(100)*4)
y <- vector("character", length=length(x))
for (i in 1:length(x)) {
   y[i] <- switch (x[i]+1, "Sun", "Rain", "Snow", "Hail", default=NA)
}

table(y)
```

## Removing for loops (cont'd)

```{r}
# the switch statement on the previous slide is equivalent to 
# the following lines:
for (i in 1:length(x)) {
  if (x[i] == 0) y[i] <- "Sun"
  if (x[i] == 1) y[i] <- "Rain"
  if (x[i] == 2) y[i] <- "Snow"
  if (x[i] == 3) y[i] <- "Hail"
}

# using R's vectors we can reduce this to a single line:
y <- c("Sun", "Rain", "Snow", "Hail")[x+1]
```


## Spotting bad code

- Ken Thompson: keep the number of indentations down, high number of local variables is suspicious
- Hadley Wickham: functions with more than ten lines are suspicious, highly repetitive code
- Generally Good Practices: indentation and spacing, comments, clear naming convention

## Profiling

Code can be profiled at different levels:

1. two (or more) alternatives of code with the same objectives are timed to determined the one which is the fastest (```system.time()``` and ```proc.time()```)
2. one piece of code is analyzed to find weak/slow spots (e.g. `profr` and `profviz`)


## Profiling process

- Collect information on code performance:
- memory usage
- frequency and duration of function calls
- Goal: optimization of code

## `system.time()`

- Probably simplest of all measures
- Evaluates the time spent in the whole expression
- doesn't give any clues about what is done at a lower level <br><br>
- Alternative: ```proc.time()```

## (Mini) code blocks

<div class="double">
<p class="double-flow">
```{r message=FALSE}
ptm <- proc.time()
t1 <- read.csv("../data/temps.csv")
proc.time() - ptm

ptm <- proc.time()
t2 <- readr::read_csv("../data/temps.csv", progress = FALSE)
proc.time() - ptm
```
</p><p class="double-flow">
```{r message=FALSE}
ptm <- proc.time()
t3 <- readRDS("../data/temps.rds")
proc.time() - ptm

ptm <- proc.time()
load("../data/temps.Rda")
proc.time() - ptm
```
</p>
</div>

## Your turn (5 min)

Which of the three options is the fastest?

```{r message=FALSE}
x <- floor(runif(100)*4)
y <- vector("character", length=length(x))

# Option #1
for (i in 1:length(x)) {
   y[i] <- switch (x[i]+1, "Sun", "Rain", "Snow", "Hail", default=NA)
}

# Option #2
for (i in 1:length(x)) {
  if (x[i] == 0) y[i] <- "Sun"
  if (x[i] == 1) y[i] <- "Rain"
  if (x[i] == 2) y[i] <- "Snow"
  if (x[i] == 3) y[i] <- "Hail"
}

# Option #3
y <- c("Sun", "Rain", "Snow", "Hail")[x+1]
```

## Your Turn Result

Based on 100 evaluations of samples of size 5000: 

```{r message=FALSE, warning=FALSE, echo = FALSE}
library(tidyverse)

results <- 1:100 %>% purrr::map_df(.f = function(x) {
x <- floor(runif(5000)*4)
y <- vector("character", length=length(x))

ptmbase <- proc.time()
# Option #1
for (i in 1:length(x)) {
   y[i] <- switch (x[i]+1, "Sun", "Rain", "Snow", "Hail", default=NA)
}
ptm1 <- proc.time()

# Option #2
for (i in 1:length(x)) {
  if (x[i] == 0) y[i] <- "Sun"
  if (x[i] == 1) y[i] <- "Rain"
  if (x[i] == 2) y[i] <- "Snow"
  if (x[i] == 3) y[i] <- "Hail"
}

ptm2 <- proc.time()

# Option #3
y <- c("Sun", "Rain", "Snow", "Hail")[x+1]

ptm3 <- proc.time()
data.frame(value=c((ptm1-ptmbase)["elapsed"], (ptm2-ptm1)["elapsed"], (ptm3-ptm2)["elapsed"]), Option=1:3)
})

results %>% 
  ggplot(aes(x = Option, y = value, fill= factor(Option))) +geom_boxplot() +
  ylab("Elapsed time")
```


## Profiling Code

Last method is much faster. But why?

`profr`/`profvis` package:

- wrapper around function ```Rprof```
- call stack is written every 0.02 seconds (interval) into a temporary file
- output file is processed to produce a summary of the usage
- time measurements varies by platform: on Unix-alike time is CPU time of the R process (excludes time waiting for input)
- Note: profiling takes time, too; once the timer goes off, the information is not recorded until the next timing click (probably in the range 1–10 msecs). Can’t go below that.


## `profvis` example

```{r eval=FALSE}
library(profvis)

p <- profvis({			
  data(diamonds, package = "ggplot2")			
  plot(price ~ carat, data = diamonds)			
  m <- lm(price ~ carat, data = diamonds)			
  abline(m, col = "red")			
})
```

`profvis` creates interactive response with code and flame plot.

Copy and paste the code into your console to run it.


